---
title: "catgraph vs related packages: a structured comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{catgraph vs related packages: a structured comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)

has_nhanes      <- requireNamespace("NHANES",      quietly = TRUE)
has_linkspotter <- requireNamespace("linkspotter",  quietly = TRUE)
has_corrgrapher <- requireNamespace("corrgrapher",  quietly = TRUE)
has_nimaa       <- requireNamespace("NIMAA",        quietly = TRUE)
has_factominer  <- requireNamespace("FactoMineR",   quietly = TRUE)
has_factoextra  <- requireNamespace("factoextra",   quietly = TRUE)
```

## Why this vignette

`catgraph` is not the only R package that visualises associations between
categorical variables as a network. This vignette places it in context by
running four approaches on the same dataset — the NHANES adult health survey —
so readers can see directly what each tool says, where `catgraph` adds
information the alternatives do not provide, and where the alternatives have
advantages worth knowing about.

The four approaches compared:

1. **`catgraph`** — all-pairwise nominal association network with four
   paradigms (Cramér's V, NMI, AMI, Bayesian Cramér's V), modality-level
   co-association network, gravity indices, and permutation tests.
2. **`linkspotter`** — computes and visualises bivariate correlations for
   mixed-type data (numeric + categorical), using NMI for categorical pairs.
3. **`corrgrapher`** — association network for mixed-type data using
   chi-square p-value transformations for categorical pairs.
4. **`NIMAA`** — bipartite graph-based mining of two-column nominal datasets,
   with clustering of projected unipartite networks.

All four are descriptive tools. None estimates conditional independence or
causal structure. The comparison is about scope, depth of analysis, and
appropriateness for all-nominal datasets.

---

## Data: NHANES adult health survey

The `NHANES` package provides a 10,000-row subset of the US National Health
and Nutrition Examination Survey. We select twelve nominal categorical
variables covering health status, behaviours, and sociodemographics, and
restrict to adults (Age >= 20) to avoid child-specific missing patterns.

```{r data, eval = has_nhanes}
library(NHANES)

nhanes_cat <- NHANES[NHANES$Age >= 20, c(
  "Gender",        # 2 levels: female, male
  "AgeDecade",     # 5 levels for adults: 20s through 60+
  "Race1",         # 5 levels: Black, Hispanic, Mexican, White, Other
  "Education",     # 5 levels: 8th Grade through College Grad
  "MaritalStatus", # 6 levels
  "BMI_WHO",       # 4 levels: underweight, normal, overweight, obese
  "HealthGen",     # 5 levels: Excellent through Poor
  "Diabetes",      # 2 levels: No, Yes
  "PhysActive",    # 2 levels: No, Yes
  "SleepTrouble",  # 2 levels: No, Yes
  "Depressed",     # 3 levels: None, Several, Most
  "Smoke100"       # 2 levels: No, Yes (ever smoked 100+ cigarettes)
)]

nhanes_cat <- droplevels(nhanes_cat)

cat("Dimensions:", nrow(nhanes_cat), "x", ncol(nhanes_cat), "\n")
cat("Variable pairs:", choose(ncol(nhanes_cat), 2), "\n\n")

# Missing value summary
na_counts <- colSums(is.na(nhanes_cat))
print(na_counts)
```

The dataset has `r if(has_nhanes) nrow(nhanes_cat) else "~6,000"` adult rows,
12 variables, and 66 variable pairs. Missing values occur because NHANES uses
age-restricted questionnaires — pairwise deletion handles this transparently
in `catgraph`.

---

## 1. catgraph

### Variable-level network — four paradigms

`catgraph` builds an all-pairwise variable-level association network. The
`method` argument selects the association paradigm; all four are shown here
for comparison.

```{r catgraph-variable, eval = has_nhanes}
library(catgraph)

# Build all four paradigm graphs
cg_cv  <- suppressWarnings(catgraph(nhanes_cat, method = "cramers_v_corrected"))
cg_nmi <- suppressWarnings(catgraph(nhanes_cat, method = "nmi"))
cg_ami <- suppressWarnings(catgraph(nhanes_cat, method = "ami"))
cg_bay <- suppressWarnings(catgraph(nhanes_cat,
                                     method = "bayesian_cramers_v",
                                     alpha  = 0.5))

# Print the Cramér's V result
cg_cv
```

```{r catgraph-summary, eval = has_nhanes}
summary(cg_cv, top = 10)
```

```{r catgraph-plot, eval = has_nhanes, fig.width = 7, fig.height = 5}
# Prune to significant edges and plot
cg_p <- prune_edges(cg_cv, min_weight = 0.05, max_p = 0.05, p_adjust = "BH")
plot(cg_p, layout = "fr", title = "NHANES — Cramér's V (bias-corrected)")
```

### Multi-paradigm rank agreement

One of `catgraph`'s unique features is the ability to assess robustness of
the association structure across paradigms.

```{r catgraph-paradigm-agree, eval = has_nhanes}
# Dense similarity matrices for rank comparison
S_cv  <- suppressWarnings(
  assoc_similarity(nhanes_cat, method = "cramers_v_corrected"))
S_nmi <- suppressWarnings(
  assoc_similarity(nhanes_cat, method = "nmi"))
S_ami <- suppressWarnings(
  assoc_similarity(nhanes_cat, method = "ami"))
S_bay <- suppressWarnings(
  assoc_similarity(nhanes_cat, method = "bayesian_cramers_v"))

ij <- which(upper.tri(S_cv), arr.ind = TRUE)

rho_cv_nmi <- cor(S_cv[ij],  S_nmi[ij], method = "spearman")
rho_cv_ami <- cor(S_cv[ij],  S_ami[ij], method = "spearman")
rho_cv_bay <- cor(S_cv[ij],  S_bay[ij], method = "spearman")
rho_nmi_ami <- cor(S_nmi[ij], S_ami[ij], method = "spearman")

cat("Spearman rank correlations between paradigms:\n")
cat(sprintf("  CV  vs NMI : %.3f\n", rho_cv_nmi))
cat(sprintf("  CV  vs AMI : %.3f\n", rho_cv_ami))
cat(sprintf("  CV  vs BAY : %.3f\n", rho_cv_bay))
cat(sprintf("  NMI vs AMI : %.3f\n", rho_nmi_ami))
cat("\nHigh rho (> 0.90) = structural conclusions are robust to paradigm choice.\n")
```

### Heatmap

```{r catgraph-heatmap, eval = has_nhanes}
plot_heatmap(cg_cv,
             show_sig = TRUE,
             title    = "NHANES — pairwise association heatmap")
```

### Modality-level network

```{r catgraph-modality, eval = has_nhanes, fig.width = 8, fig.height = 6}
# Use Titanic for the modality graph — compact and interpretable
df_titanic <- expand_table(Titanic)

mg <- build_modality_graph(df_titanic, method = "cramers_v_corrected")
mg <- prune_modality_edges(mg, min_weight = 0.10)
mg <- cluster_modalities(mg)

plot(mg, color_by = "cluster",
     main = "Titanic modality graph — Cramér's V")
```

```{r catgraph-gravity, eval = has_nhanes}
# Gravity indices: which modalities are attractors vs satellites?
grav <- modality_gravity(mg)
print(grav)
```

### Cross-group comparison

```{r catgraph-crossgroup, eval = has_nhanes}
# Build sex-stratified modality graphs on NHANES
mg_f <- suppressWarnings(
  build_conditional_modality_graph(nhanes_cat,
                                    given = list(Gender = "female")))
mg_m <- suppressWarnings(
  build_conditional_modality_graph(nhanes_cat,
                                    given = list(Gender = "male")))

mg_f <- prune_modality_edges(mg_f, min_weight = 0.08, max_p = 0.05)
mg_m <- prune_modality_edges(mg_m, min_weight = 0.08, max_p = 0.05)

# Permutation test: do female and male networks differ structurally?
set.seed(42)
perm_test <- test_modality_graph_equality(mg_f, mg_m, n_perm = 200)
print(perm_test)
```

**What `catgraph` provides that no other package in this comparison does:**

- All-pairwise variable association network for all-nominal datasets
- Four association paradigms with documented scale differences
- Modality-level co-association network (category levels as nodes)
- Modality gravity indices (MGI, dMGI, Orbital Score)
- Sex-stratified or group-stratified conditional modality graphs
- Permutation-based test for cross-group structural differences
- Bootstrap confidence intervals on edge weights

---

## 2. linkspotter

`linkspotter` computes and visualises bivariate associations for
**mixed-type** datasets (numeric + categorical). For categorical-categorical
pairs it uses a discretisation-based normalised mutual information (MaxNMI).
Its strength is handling mixed data in a single unified network; its
limitation for an all-nominal dataset is that it provides no modality-level
representation and no multi-paradigm comparison.

```{r linkspotter-run, eval = has_linkspotter && has_nhanes}
library(linkspotter)

# linkspotter requires at least some numeric columns; for a pure categorical
# dataset we demonstrate on a small subset to show the interface
nhanes_small <- nhanes_cat[, c("Gender","HealthGen","Diabetes",
                                "PhysActive","Smoke100")]
nhanes_small <- nhanes_small[complete.cases(nhanes_small), ]

# Compute all bivariate links
cor_couples <- suppressWarnings(
  multiBivariateCorrelation(nhanes_small)
)

# Show the result table
head(cor_couples[order(cor_couples$MaxNMI, decreasing = TRUE), ], 10)
```

```{r linkspotter-plot, eval = has_linkspotter && has_nhanes,
    fig.width = 7, fig.height = 5}
# Network plot using MaxNMI
linkspotterGraph(
  corDF             = cor_couples,
  corMethod         = "MaxNMI",
  minCor            = 0.05,
  smoothEdges       = FALSE,
  dynamicNodes      = FALSE
)
```

**Honest assessment of linkspotter vs catgraph for nominal data:**

| Feature | linkspotter | catgraph |
|---|---|---|
| Mixed numeric + categorical | ✔ best-in-class | ✗ nominal only |
| All-nominal datasets | ✔ works | ✔ native design |
| Association metric choice | Fixed by variable type | 4 paradigms, user-selected |
| Bias correction for Cramér's V | ✗ | ✔ |
| Bayesian edge weights | ✗ | ✔ |
| Modality-level network | ✗ | ✔ |
| Gravity indices | ✗ | ✔ |
| Cross-group permutation test | ✗ | ✔ |
| Interactive Shiny app | ✔ | ✗ |

**Use `linkspotter` when** your dataset mixes numeric and categorical columns
and you want a quick unified overview. **Use `catgraph` when** all variables
are nominal and you need multi-paradigm robustness checking, modality-level
structure, or cross-group inference.

---

## 3. corrgrapher

`corrgrapher` builds an association network for mixed-type data. For two
categorical variables it uses the chi-square test p-value transformed to a
[0, 1] association scale: `assoc = 1 - p_value`. This is a very crude measure
for categorical pairs — it confounds sample size with effect size, meaning a
weak association observed in a large sample can produce a higher edge weight
than a strong association in a smaller subsample.

```{r corrgrapher-run, eval = has_corrgrapher && has_nhanes}
library(corrgrapher)

nhanes_small2 <- nhanes_cat[, c("Gender","HealthGen","Diabetes",
                                 "PhysActive","Smoke100","Depressed")]
nhanes_small2 <- nhanes_small2[complete.cases(nhanes_small2), ]

# corrgrapher handles factors natively
# print() and plot() open interactive widgets — inspect the object class
# and names instead, which is safe in non-interactive environments
cgr <- corrgrapher(nhanes_small2)
cat("corrgrapher object components:", paste(names(cgr), collapse = ", "), "\n")
cat("Number of nodes:", nrow(cgr$nodes), "\n")
cat("Number of edges:", nrow(cgr$edges_df), "\n\n")

# Show edge weights — corrgrapher stores them in edges_df
e <- cgr$edges_df
if (!is.null(e) && nrow(e) > 0) {
  e <- e[order(e$value, decreasing = TRUE), ]
  cat("Top associations by weight:\n")
  print(head(e[, c("from","to","value")], 10), row.names = FALSE)
}
```

```{r corrgrapher-plot, eval = FALSE}
# plot(cgr) opens an interactive visNetwork widget — run interactively only
plot(cgr)
```

**Honest assessment of corrgrapher vs catgraph for nominal data:**

| Feature | corrgrapher | catgraph |
|---|---|---|
| Mixed numeric + categorical | ✔ | ✗ nominal only |
| Edge weight for categorical pairs | 1 − p-value (confounds n and effect) | Effect size in [0,1] |
| Bias correction | ✗ | ✔ |
| Multiple paradigms | ✗ | ✔ (4 paradigms) |
| Modality-level network | ✗ | ✔ |
| Gravity indices | ✗ | ✔ |
| Integrated with `DALEX` explainers | ✔ | ✗ |

**Key methodological difference:** `corrgrapher`'s p-value-based edge weights
are sample-size dependent. On NHANES (n ≈ 6,000), nearly all pairs will have
p < 0.05, making the weights cluster near 1 and providing little
discriminative power. `catgraph`'s effect-size-based weights remain
informative regardless of sample size.

---

## 4. NIMAA

`NIMAA` takes a fundamentally different approach. Rather than building an
all-pairwise variable network, it takes **two nominal columns** and constructs
a **bipartite graph** between the levels of those two variables. It then
projects this bipartite graph onto each variable's levels to obtain unipartite
networks, and applies clustering to find groups of similar category labels.

This makes NIMAA ideal for problems like "which drug labels cluster together
based on their co-occurrence with patient labels?" — a two-variable bipartite
mining task. It is not designed for all-pairwise multi-variable association
networks.

```{r nimaa-run, eval = has_nimaa && has_nhanes}
library(NIMAA)

# NIMAA requires a two-column nominal input (optionally with a weight column)
# We demonstrate on HealthGen x Depressed — two health status variables
nimaa_input <- nhanes_cat[, c("HealthGen", "Depressed")]
nimaa_input <- nimaa_input[complete.cases(nimaa_input), ]

# Convert to bipartite network incidence matrix
binet <- nominalAsBinet(nimaa_input)
cat("Incidence matrix dimensions:", dim(binet), "\n")
print(binet)
```

```{r nimaa-plot, eval = has_nimaa && has_nhanes,
    fig.width = 7, fig.height = 5}
# Plot the bipartite network and its projections
plotBipartite(binet, vertex.label.cex = 0.8)
```

```{r nimaa-cluster, eval = has_nimaa && has_nhanes}
# Inspect the incidence matrix returned by nominalAsBinet()
# (findCluster() requires a plain matrix — extract it if binet is a list)
inc_mat <- if (is.list(binet)) binet[[1]] else binet

cat("Incidence matrix class:", class(inc_mat), "\n")
cat("Dimensions:", nrow(inc_mat), "x", ncol(inc_mat), "\n\n")
cat("Row names (HealthGen levels):", rownames(inc_mat), "\n")
cat("Col names (Depressed levels):", colnames(inc_mat), "\n\n")

# Show the co-occurrence counts
cat("Co-occurrence matrix (HealthGen x Depressed):\n")
print(inc_mat)
```

**Honest assessment of NIMAA vs catgraph for nominal data:**

| Feature | NIMAA | catgraph |
|---|---|---|
| Number of variables handled simultaneously | 2 | All pairwise (p variables) |
| Network type | Bipartite (respondent × modality style) | Unipartite variable network + unipartite modality network |
| Edge weight | Co-occurrence count or optional numeric | Effect size (4 paradigms) |
| Missing value imputation | ✔ (unique feature) | ✗ (pairwise deletion only) |
| Clustering of category labels | ✔ (primary purpose) | ✔ (via `cluster_modalities()`) |
| All-variable pairwise summary | ✗ | ✔ |
| Gravity indices | ✗ | ✔ |
| Cross-group permutation test | ✗ | ✔ |
| Biological / bioinformatics focus | ✔ | ✗ (social science / survey focus) |

**Key distinction:** NIMAA's bipartite approach is the right tool when you have
exactly two nominal variables and want to cluster their levels based on
co-occurrence patterns. `catgraph`'s modality graph answers a similar question
for all variable pairs simultaneously, with statistical filtering via
effect-size thresholds and p-value correction.

---

## Summary comparison

```{r summary-table, eval = TRUE}
comparison <- data.frame(
  Feature = c(
    "All-pairwise nominal network",
    "Mixed numeric + categorical",
    "Effect-size edge weights",
    "Multiple association paradigms",
    "Bias-corrected Cramér's V",
    "Bayesian edge weights",
    "Modality / category-level network",
    "Gravity indices (MGI, dMGI)",
    "Cross-group permutation test",
    "Bootstrap CIs on edges",
    "Missing value imputation",
    "Interactive Shiny app",
    "Bipartite (2-variable) mining",
    "DALEX integration"
  ),
  catgraph     = c("✔","✗","✔","✔","✔","✔","✔","✔","✔","✔","✗","✗","✗","✗"),
  linkspotter  = c("✔","✔","✗","✗","✗","✗","✗","✗","✗","✗","✗","✔","✗","✗"),
  corrgrapher  = c("✔","✔","✗","✗","✗","✗","✗","✗","✗","✗","✗","✗","✗","✔"),
  NIMAA        = c("✗","✗","✗","✗","✗","✗","✔","✗","✗","✗","✔","✗","✔","✗"),
  stringsAsFactors = FALSE
)

knitr::kable(comparison,
             col.names = c("Feature","catgraph","linkspotter",
                           "corrgrapher","NIMAA"),
             align = "lcccc")
```

### When to use which package

**`catgraph`** — all variables are nominal; you want a multi-variable
association network with statistically grounded edge weights, modality-level
structure, gravity indices, or cross-group inference.

**`linkspotter`** — your dataset mixes numeric and categorical columns and you
want a quick interactive overview of all bivariate links in a single network.

**`corrgrapher`** — you are already in the `DALEX` model-explanation ecosystem
and want a network of feature associations alongside model-level explanations.

**`NIMAA`** — you have exactly two nominal variables and want to cluster their
category labels based on co-occurrence, or you need missing-value imputation
for a bipartite nominal network.

---

## References

- Csardi, G., & Nepusz, T. (2006). The igraph software package for complex
  network research. *InterJournal, Complex Systems*, 1695.
- Jafari, M., et al. (2022). NIMAA: an R/CRAN package to accomplish NomInal
  data Mining AnAlysis. *bioRxiv*. <https://doi.org/10.1101/2022.01.13.475835>
- Johnson, R. W. (2020). NHANES: Data from the US National Health and
  Nutrition Examination Survey. R package.
  <https://CRAN.R-project.org/package=NHANES>
- Morgen, K., & Biecek, P. (2020). corrgrapher: Explore correlations between
  variables in a machine learning model. R package.
  <https://CRAN.R-project.org/package=corrgrapher>
- Samba, A. (2020). linkspotter: Bivariate correlations calculation and
  visualization. R package.
  <https://CRAN.R-project.org/package=linkspotter>
- Vinh, N. X., Epps, J., & Bailey, J. (2010). Information theoretic measures
  for clusterings comparison. *Journal of Machine Learning Research*, 11,
  2837–2854.
