---
title: "Comparing catgraph modality networks with MCA and bipartite approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing catgraph modality networks with MCA and bipartite approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)

has_factominer  <- requireNamespace("FactoMineR",  quietly = TRUE)
has_factoextra  <- requireNamespace("factoextra",  quietly = TRUE)
have_mca_stack  <- has_factominer && has_factoextra
```

## Why this vignette

`catgraph`'s modality layer is one of several established traditions for
analysing the structure of categorical data. This vignette runs the
modality-level workflow on the bundled `survey_health` dataset alongside
three alternative methods, so readers can see directly what each method
says — and, crucially, where `catgraph` adds information and where it
should **not** be overclaimed. A final section introduces
`joint_balance()`, the high-level cross-group diagnostic shipped in
v0.8.0.

The methods compared:

1. **`catgraph` modality network** — chi-square / phi-filtered
   co-association graph on modalities. Scope: which category levels
   significantly co-associate across variables.
2. **Multiple Correspondence Analysis (MCA)** — geometric approach.
   Scope: low-dimensional embedding of modalities in a continuous space
   based on chi-square distances between row and column profiles.
3. **Bipartite respondent–modality network** — two-mode incidence graph.
   Scope: the raw unprojected endorsement structure, before any
   statistical filter.
4. **Naive co-occurrence projection** — the bipartite graph projected
   onto its modality partition, with edges weighted by raw count of
   shared respondents. Scope: which modality pairs are most
   frequently endorsed together, without adjustment for base rates.

All four are descriptive methods. None is a substitute for the others;
each answers a different question about the same categorical data.

## Setup

```{r load}
library(catgraph)
library(igraph)

data(survey_health)
str(survey_health)
```

`survey_health` is a synthetic dataset with 600 respondents, eight
categorical variables, and known designed-in association structure.
See `?survey_health` for details.

## Method 1 — catgraph modality network

```{r catgraph-mod}
mg <- build_modality_graph(survey_health)
mg <- prune_modality_edges(mg, min_weight = 0.10, max_p = 0.05)
set.seed(1)
mg <- cluster_modalities(mg, method = "louvain")

mg
```

The pruned, clustered modality network is our primary object of
interest. Plot it with `signed = TRUE` to reveal the direction of
association encoded in the stored standardised Pearson residuals:

```{r catgraph-mod-plot, fig.height = 6}
plot(mg, color_by = "cluster", signed = TRUE)
```

Green edges indicate modalities that co-occur more than expected under
independence (attraction); red edges indicate co-occurrence below
expectation (repulsion). Communities that mix green and red edges are
showing modalities that cluster by the *magnitude* of their chi-square
signal, which can conflate opposing substantive patterns — a known
limitation of unsigned Louvain on a phi-weighted graph.

Summarise the community composition:

```{r catgraph-mod-summary}
comm <- summarise_modality_communities(mg)
comm$community_summary
```

Communities here should be interpreted as **groups of co-associated
modalities**. They are not respondent segments and they are not a
low-dimensional embedding.

## Method 2 — Multiple Correspondence Analysis (MCA)

MCA is the classical geometric approach to categorical multivariate
structure. It projects both rows (respondents) and columns (modalities)
onto a shared low-dimensional Euclidean space where proximity reflects
association strength.

```{r mca, eval = have_mca_stack, fig.height = 6}
library(FactoMineR)
library(factoextra)

mca_df <- stats::na.omit(survey_health)
mca_df[] <- lapply(mca_df, function(x) if (is.factor(x)) x else as.factor(x))

mca_res <- MCA(mca_df, graph = FALSE)

fviz_mca_var(
  mca_res, repel = TRUE,
  col.var       = "contrib",
  gradient.cols = c("#AFA9EC", "#7F77DD", "#534AB7"),
  ggtheme       = ggplot2::theme_minimal(base_size = 10),
  title         = "MCA biplot: modality map (first two dimensions)"
)
```

```{r mca-fallback, eval = !have_mca_stack, echo = FALSE, results = "asis"}
cat(
  "> *The MCA block requires `FactoMineR` and `factoextra`, which are ",
  "listed in `Suggests:`. Install them with `install.packages(c(",
  "'FactoMineR', 'factoextra'))` to see the biplot.*\n"
)
```

What the MCA biplot shows, that the modality graph does not:

* **Continuous geometry.** Two modalities plotted far from the origin on
  the same axis are strongly associated with that axis, which summarises
  the dominant pattern of variation in the data. Modalities near the
  origin are "average" — they don't load heavily on any dimension.
* **Contribution colour scale.** The `contrib` colouring shows how much
  each modality contributes to the first two MCA dimensions; a bright
  modality is driving the overall structure, a pale one is passive.
* **Row-column duality.** MCA would also place respondents on the same
  map (not shown above), enabling direct respondent–modality comparison
  in a single coordinate system.

What the MCA biplot does not show, that the modality graph does:

* **Discrete edges with a significance filter.** MCA gives every
  modality a coordinate regardless of whether its associations with
  other modalities are statistically significant.
* **Community structure with an algorithmic cluster assignment.** MCA
  produces a geometry; communities must be imposed on top via
  `FactoMineR::HCPC()` or similar.
* **Signed pairwise information.** MCA coordinates encode a complex
  blend of many associations; you cannot read "these two modalities
  attract, those two repel" off the biplot directly.

**Bottom line:** MCA is the right tool if you want a low-dimensional
continuous map of both respondents and categories. The modality graph
is the right tool if you want a discrete, filtered, pairwise picture
with explicit communities.

## Method 3 — Bipartite respondent–modality network

The bipartite graph preserves the raw incidence structure: respondents
on one side, modalities on the other, an edge whenever a respondent
endorses a modality. Nothing is filtered or projected.

```{r bipartite, fig.height = 6}
bg <- bipartite_modality_graph(survey_health)
bg
```

```{r bipartite-plot, fig.height = 6}
plot(bg, max_respondents = 150)
```

The bipartite view is honest about what categorical survey data actually
is: a set of respondents and a set of modalities they endorsed. Every
other method in this vignette is derived from this bipartite structure
via some kind of projection or statistical filter.

**When to use the bipartite graph directly:**

* Affiliation-network analysis (persons to events, authors to papers,
  species to sites) — where the bipartite structure *is* the scientific
  object, not a scaffolding for something else.
* Auditing or diagnostic work — seeing which respondents endorse which
  modalities without any modelling assumption.

**When not to use it directly:**

* For substantive interpretation of modality relationships. Without
  projection and filtering, the bipartite view is too granular to
  summarise category-level structure.

## Method 4 — Naive co-occurrence projection

Projecting the bipartite graph onto its modality partition yields a
count-based co-occurrence network. Every pair of modalities gets an
edge weighted by the number of respondents who endorsed both.

```{r cooccurrence, fig.height = 6}
proj <- igraph::bipartite_projection(bg$graph, which = "true")
proj
```

Plot only the strongest 40% of projected edges, to keep the figure
readable:

```{r cooccurrence-plot, fig.height = 6}
w_proj <- igraph::E(proj)$weight
q60    <- stats::quantile(w_proj, 0.60, na.rm = TRUE)
proj_strong <- igraph::delete_edges(proj, which(w_proj < q60))
proj_strong <- igraph::delete_vertices(
  proj_strong, which(igraph::degree(proj_strong) == 0)
)

mod_var <- sub("=.*", "", igraph::V(proj_strong)$name)
pal <- grDevices::hcl.colors(length(unique(mod_var)), palette = "Dark 3")
v_col <- pal[as.integer(factor(mod_var))]

op <- graphics::par(mar = c(1, 1, 3, 1))
plot(
  proj_strong,
  layout           = igraph::layout_with_fr(proj_strong),
  vertex.size      = 10,
  vertex.color     = v_col,
  vertex.label.cex = 0.65,
  vertex.label.color = "#2C2C2A",
  edge.width       = 0.3,
  edge.color       = grDevices::adjustcolor("grey40", alpha.f = 0.3),
  main = "Naive co-occurrence projection (top 40% edges)"
)
graphics::par(op)
```

This is the key comparison with the `catgraph` modality network. Both
are modality-only graphs. Both have edges between modalities. But they
are weighted differently:

* **Co-occurrence projection:** edge weight = raw count of shared
  respondents. High-frequency modalities appear central because they
  are endorsed by many respondents, not because they are substantively
  associated with their neighbours.
* **`catgraph` modality network:** edge weight = phi on the 2×2
  indicator table, after chi-square significance filtering. Adjusts
  for base rates — a high-frequency modality only gets a strong edge
  to a neighbour if their co-occurrence exceeds what their marginal
  frequencies would predict under independence.

The difference matters. If you are comparing populations (e.g. two
countries or two ecological sites), raw co-occurrence will always
emphasise whatever modalities are most common in each population,
which can obscure substantive structural differences. `catgraph` is
designed to filter out base-rate effects so you see the signal.

## Summary table: when to use what

| Question you have                                                     | Method                                              |
|-----------------------------------------------------------------------|-----------------------------------------------------|
| Which category levels *significantly* co-associate across variables?  | **`catgraph` modality network**                     |
| What is the low-dimensional geometry of my categories?                | **MCA** (`FactoMineR::MCA`)                         |
| Who endorsed what, unfiltered?                                        | **Bipartite graph** (`bipartite_modality_graph`)    |
| Which modalities are most often endorsed together, raw?               | **Co-occurrence projection**                        |
| Are two groups *jointly* balanced on their categorical structure?     | **`joint_balance()`** (see next section)            |
| Who are my respondent segments?                                       | `poLCA`, `FactoMineR::HCPC()`                       |
| Do my variables show conditional-independence structure?              | `bnlearn`, `gRim`, or GGM packages                  |

## What `catgraph` adds

1. **Statistical filtering native to nominal data.** Unlike MCA (which
   embeds everything) or co-occurrence projection (which filters only
   by count), `catgraph` uses chi-square significance plus Bergsma-
   corrected effect sizes to decide which associations are substantive.
2. **Explicit communities with a clustering algorithm.** Louvain (or
   walktrap) on the phi-weighted graph gives discrete community
   assignments that MCA does not provide without an additional
   hierarchical-clustering step.
3. **Signed interpretation via stored residuals.** The
   `std_resid` edge attribute retains the direction of association,
   which is lost in both MCA coordinates and unsigned co-occurrence
   counts.
4. **A shared framework for variable-level and modality-level
   analysis.** `catgraph` and `build_modality_graph` produce objects
   with consistent conventions (pruning, community detection,
   visualisation), making it natural to move between the two layers.
5. **A cross-group diagnostic workflow.** `joint_balance()` uses the
   modality-graph machinery to assess whether two or more subgroups
   share the same joint categorical structure, which no purely
   geometric or purely count-based method does directly.

## Where `catgraph` should not be overclaimed

* It is **not** a respondent-segmentation tool. Communities are groups
  of modalities, not groups of respondents. Use `poLCA` or
  `FactoMineR::HCPC()` for segmentation.
* It is **not** an MCA replacement. If you want a geometry, use MCA.
* It is **not** a conditional-independence model. Edges encode marginal
  pairwise dependence only. Use graphical models for conditional
  structure.
* Its unsigned phi weights can mix attraction with repulsion in the
  same community. Interpret cluster structure alongside the signed
  residual view (`signed = TRUE` in `plot.catmodgraph`).

# Joint balance diagnostics for grouped categorical data

## The problem

Suppose you are combining two samples — two survey waves, two study
sites, two arms of a meta-analysis, or two levels of a sociodemographic
stratum. Standard practice is to report a "Table 1": percentages of
each categorical variable broken down by group, with a chi-square test
per variable, so the reader can see whether the groups are comparable.

This is a *marginal* check. It asks, for each variable separately:
is the distribution of that variable the same across groups? It does
**not** ask whether the *joint* distribution is the same. Two groups
can pass every marginal check and still differ in joint structure —
for example, if the association between education and political
orientation has one sign in one group and the opposite sign in
another, the marginal distributions of education and of political
orientation can be identical across groups while the joint
distribution differs substantially. A marginal-only diagnostic
misses this.

For meta-analysis, repeated cross-sectional surveys, and
multi-site observational studies, this is not an abstract concern.
It is the reason heterogeneity is often diagnosed by marginal
indicators, pooled by simple weighting, and only later found to
have been driven by differences in the joint covariate structure
that the marginal-check step could not detect.

## The diagnostic

`joint_balance()` is a complete three-step diagnostic for this
situation. It takes a categorical data frame plus a grouping
variable and runs:

1. **Marginal tests.** A chi-square test per variable, with
   Benjamini-Hochberg adjustment across variables. This is the
   classical Table 1 check, reported for continuity with prior
   practice.

2. **Pairwise omnibus tests.** For each pair of groups, the
   permutation test from `test_modality_graph_equality()` is run
   on the modality-level association network. This tests whether
   the full joint categorical structure is the same across the
   pair. Bonferroni adjustment is applied across the group pairs.

3. **Edge-wise post-hoc tests.** For each group pair whose omnibus
   test rejects, `test_modality_edge_differences()` identifies
   which specific modality-pair associations drive the difference,
   with Benjamini-Hochberg adjustment across edges within the
   pair.

The three steps have three separate multiple-testing families.
This is deliberate: they answer three different questions and
should not be bundled into a single correction.

```{r joint-balance-example, eval = FALSE}
data(survey_health)

jb <- joint_balance(
  survey_health,
  group       = "sex",
  n_perm      = 500,
  n_perm_edge = 500,
  seed        = 1
)

jb
summary(jb)
plot(jb)
```

The `print()` method gives a three-block summary: how many marginal
variables rejected, how many group pairs rejected, and within each
rejecting pair how many edges differ significantly. The `plot()`
method produces a two-panel diagnostic: a bar chart of adjusted
marginal p-values on the left, and the modality-difference graph
for the most-rejecting pair on the right.

The chunk above is set `eval = FALSE` because a 500-permutation run
inflates vignette build time noticeably; the output is identical in
form to the print methods shown in the function's `?joint_balance`
examples.

## When to use this instead of the alternatives

`joint_balance()` is not the only way to check cross-group
comparability of categorical data. Three established alternatives
are worth naming explicitly, because each is the better tool for
part of this space.

### Log-linear homogeneity-of-association tests

The textbook tool for "does the A–B association differ across
groups?" is the log-linear model of the three-way table, with a
likelihood-ratio test of the no-three-way-interaction null
(Agresti, *Categorical Data Analysis*, chapter 9). For a single
pair of variables this is more powerful than the permutation test
used here, more interpretable because the model gives
log-odds-ratio differences with standard errors, and exact under
large-sample theory.

**Use the log-linear approach when:** you have a specific
two-variable association hypothesis to test; the three-way table
is well-populated (no sparse cells); the cross-group comparison
is the primary scientific question and the analysis is
confirmatory.

**Use `joint_balance()` when:** you have many variables (say, more
than four), the log-linear model would require specifying and
comparing a combinatorially large set of interaction terms; the
analysis is exploratory, diagnostic, or descriptive; the purpose
is to decide whether the groups are comparable enough to pool,
rather than to characterise the specific form of any difference.

These are complementary roles. A reasonable workflow uses
`joint_balance()` as a first-pass screen, then fits a log-linear
model to any specific association the edge-wise step flags as
differing.

### Multiple Correspondence Analysis with supplementary group

MCA with the grouping variable as a supplementary categorical
projects the groups onto the modality coordinate space, so the
reader can see whether groups occupy distinguishable regions of
the joint category geometry (see Greenacre & Blasius,
*Multiple Correspondence Analysis and Related Methods*, 2006;
implemented as `FactoMineR::MCA(..., quali.sup = ...)`).

**Use MCA when:** the goal is a global visual summary of the
category space with the groups located within it; the
dimensionality of the joint structure is itself of interest; you
are presenting a single low-dimensional figure to describe
between-group differences.

**Use `joint_balance()` when:** the goal is a pass/fail
comparability verdict with edge-level localisation, rather than a
low-dimensional representation; you want a formal p-value rather
than a visual judgement; the reader needs to know *which*
specific modality pairs differ, not just that the groups land in
different regions of a projection.

Again, these compose. The MCA plot is often the right figure to
accompany a significant `joint_balance()` result, with the MCA
showing the overall geometric separation and the
`plot_modality_difference()` graph showing the edges responsible.

### tableone and cobalt

For the marginal Table 1 itself, `tableone::CreateTableOne()` and
`cobalt::bal.tab()` are more mature than the marginal panel of
`joint_balance()`. They handle continuous covariates (which
`joint_balance()` explicitly does not), report standardised mean
differences, and produce publication-ready formatting. The
marginal step in `joint_balance()` exists only so the reader can
see the marginal and joint views side-by-side; it is not
intended to replace these established tools.

**The recommended pattern** is to use `tableone` or `cobalt` for
the formal Table 1 in the paper, and `joint_balance()` for the
joint-structure diagnostic, citing both. The joint step is
stronger than the marginal step (joint balance implies marginal
balance) but the marginal step is stronger on continuous
variables and on reporting conventions.

## What joint balance does not do

Three limitations to state plainly, because reviewers will ask.

**It is not a causal adjustment tool.** Rejecting the joint
omnibus means the groups differ in joint categorical structure;
it does not say whether that difference confounds the outcome of
interest. Propensity-score weighting, entropy balancing, and
matching are separate methodologies and remain necessary when
the question is causal.

**It is descriptive, not model-based.** The edge weights are
bivariate phi coefficients, not conditional log-odds-ratios from
a fitted model. If you need confidence intervals on specific
three-way interaction terms, fit the log-linear model. The
permutation test gives an honest p-value for "does the overall
joint structure differ?" but the answer is global, not
parametric.

**It does not handle repeated measures.** The permutation null
assumes exchangeable rows under group relabelling. This is
satisfied by independent cross-sections (different respondents
at each wave) but violated by panel data (same respondents
followed over time). For repeated-measures comparisons, the
permutation machinery here is not the right tool; standard
longitudinal categorical methods (GEE with a working
independence structure, mixed-effects log-linear models) are.

## A worked example: conditional modality graphs

The comparison above is unconditional — every row of each group is
used. Sometimes the scientifically interesting question is
conditional: not "does the joint structure differ between two
groups?" but "within one subgroup, how does the association
structure vary across a secondary stratum?".

`build_conditional_modality_graph()` is a thin wrapper around
`build_modality_graph()` that subsets the data by one or more
conditioning levels and records the conditioning on the returned
object. The output is a plain `catmodgraph` and composes with
all existing methods.

```{r conditional-example, eval = FALSE}
data(survey_health)

# Within one age stratum, compare across levels of a second variable
ag_levels <- levels(survey_health$age_group)
subset_by_age <- lapply(
  ag_levels,
  function(ag) {
    build_conditional_modality_graph(
      survey_health,
      given = list(age_group = ag)
    )
  }
)
names(subset_by_age) <- ag_levels

compare_modality_graphs(
  subset_by_age,
  restrict = "common",
  signed   = TRUE
)
```

The conditioning specification is stored in
`mg$conditioning$given` and can be inspected if you need to
audit what a given panel represents. Note that the function is
designed for conditioning on *observed* stratification variables
only; conditioning on a cluster membership that was itself
estimated from the association structure reintroduces
circularity, and the function does not support it (the 0.6.0
release notes document this refocus in detail).

## References

* Agresti, A. (2013). *Categorical Data Analysis* (3rd ed.). Wiley.
* Greenacre, M. (2017). *Correspondence Analysis in Practice* (3rd ed.).
  Chapman and Hall/CRC.
* Greenacre, M., & Blasius, J. (Eds.) (2006). *Multiple Correspondence
  Analysis and Related Methods*. Chapman and Hall/CRC.
* Borgatti, S. P., & Everett, M. G. (1997). Network analysis of 2-mode
  data. *Social Networks*, 19(3), 243–269.
* Latapy, M., Magnien, C., & Del Vecchio, N. (2008). Basic notions for
  the analysis of large two-mode networks. *Social Networks*, 30(1),
  31–48.
* Vermunt, J. K., & Magidson, J. (2002). Latent class cluster analysis.
  In J. A. Hagenaars & A. L. McCutcheon (Eds.), *Applied Latent Class
  Analysis*. Cambridge University Press.
