---
title: "Introduction to catgraph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to catgraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)

library(catgraph)
data(survey_health)

cg_full <- catgraph(survey_health)
```

# Introduction

Categorical datasets are typically analysed through marginal summaries
(e.g., contingency tables) or dimension-reduction techniques such as
Multiple Correspondence Analysis (MCA). While these approaches are
useful, they do not directly expose the **pairwise association structure**
between variables or between specific category levels.

The `catgraph` package provides a network-based framework for exploring
categorical association structure at two complementary resolutions:

1. **Variable level** — nodes are variables; edges encode pairwise
   association strength via one of four paradigms: frequentist (Cramér's V),
   information-theoretic (NMI, AMI), or Bayesian (Dirichlet-smoothed
   Cramér's V).

2. **Modality level** — nodes are category levels (modalities); edges
   encode cross-variable co-occurrence structure using 2×2 association
   measures.

Both representations are **descriptive** and operate on marginal
associations. They do not estimate conditional independence or causal
structure.

# Conceptual framing

The methods implemented in `catgraph` belong to the class of
**pairwise association networks**.

- Edges quantify **marginal dependence** between variables or modalities.
- No conditioning is performed on other variables.
- As a result, edges may reflect indirect relationships.

This distinguishes `catgraph` from:

- **Graphical models** (e.g., Bayesian networks), which target
  conditional independence.
- **Latent variable models** (e.g., MCA, LCA), which summarise structure
  in reduced dimensions.

The overall workflow is:

```text
categorical data
  -> pairwise contingency structure
  -> weighted association network
  -> pruning, community detection, visualisation
```

## Scope and interpretation

### Variable-level graphs

A `catgraph` is a **pairwise marginal association network**, not a
conditional-independence graphical model. An edge between variables `A`
and `B` indicates that `A` and `B` are dependent in their bivariate
contingency table. It does **not** mean that `A` and `B` remain dependent
after controlling for all other variables.

Practical implications:

* **Triangles may reflect confounding.** If `A`, `B`, and `C` form a triangle,
  they may all be directly associated, or a common cause may induce
  pairwise dependence among them.
* **Central nodes need careful reading.** A highly central variable may
  reflect high cardinality, broad association, or common-cause structure.
  It should not be interpreted as a causal driver.
* **Communities describe co-variation, not independence blocks.**
  Louvain or Walktrap communities identify groups of variables that
  co-vary pairwise. They are not conditional-independence modules.

### Modality-level graphs

A modality graph is a **category co-association network**. Nodes are
modalities (factor levels) such as `smoking_status=current` or
`bmi_category=obese`. Edges are restricted to **cross-variable** modality
pairs — same-variable modalities are excluded by construction, because
modalities from the same variable are mutually exclusive by definition.
Edge weights use the chosen association measure on the 2×2 indicator table,
and the signed standardised Pearson residual is stored as a separate edge
attribute for interpretation.

**What this layer is:** a descriptive map of which category levels tend to
co-occur across different variables. It sits in the same methodological
family as Multiple Correspondence Analysis (MCA) and two-mode affiliation
networks — a tool for mapping the structure of category co-association.

**What this layer is not:** a respondent-segmentation method. Community
detection here groups *modalities*, not participants. For respondent-level
segmentation, use `poLCA` (latent class analysis) or `FactoMineR::MCA()`
followed by `FactoMineR::HCPC()` (MCA + hierarchical clustering of
respondents). These are complementary tools answering a different question,
and `catgraph` does not replace them.

### When to use `catgraph`

`catgraph` answers two specific descriptive questions. It is most useful
when those are the questions you have.

| Your question                                                   | Use                                            |
|-----------------------------------------------------------------|------------------------------------------------|
| How do categorical variables co-vary pairwise?                  | `catgraph` **variable-level network**          |
| Which category levels bundle together across variables?         | `catgraph` **modality-level network**          |
| What are the respondent segments in my data?                    | `poLCA` or `FactoMineR::HCPC()`                |
| What is the low-dimensional geometry of my categories?          | `FactoMineR::MCA()`                            |
| Who endorses what, without any modelling?                       | Bipartite affiliation graph                    |
| Are there conditional-independence relationships?               | `bnlearn`, `gRim`, or GGM packages             |

### Do **not** use `catgraph` for

* Causal inference or claims about direct effects.
* Conditional-independence structure (use a graphical model instead).
* Respondent segmentation or latent-class analysis.
* Exact equivalence to MCA geometry.

---

## Methodological caveats

Five caveats apply to every variable-level `catgraph` analysis.

**1. Marginal, not conditional.**
Edges quantify bivariate association only.

**2. Mixed metrics across table dimensions and paradigms.**
Phi (2×2) and Cramér's V (R×C) both lie in [0, 1] but are not strictly
identical in statistical behaviour across table sizes, sparsity, and marginal
imbalance. NMI and AMI weights are information-theoretic quantities, not
effect sizes, and are typically three to five times smaller in magnitude than
Cramér's V on the same data. Bayesian Cramér's V converges to classical
Cramér's V as n → ∞ but diverges on sparse cells. **Edge weights from
different paradigms are not directly comparable across graph objects.**

A useful diagnostic is to inspect both table dimensions and the metric used:

```{r caveat2-dim, eval = FALSE}
tidy <- assoc_matrix(cg_full, format = "tidy")
table(tidy$type)   # 2x2 vs RxC breakdown
```

When the graph mixes many 2×2 and many R×C edges, centrality values are only
approximately comparable across variable pairs. High-cardinality variables may
dominate centrality rankings without being substantively more associated.
Bias correction and the Bayesian estimator both mitigate, but do not fully
remove, this issue.

**3. Pairwise deletion yields edge-specific sample sizes.**
Each edge is estimated on the observations where both variables are
non-missing. If missingness is not completely at random, graph topology
may partly reflect missingness structure rather than substantive structure.

```{r n-variation, eval = FALSE}
ns <- igraph::E(cg_full$graph)$n
range(ns)
```

**4. Multiple testing.**
With `p` variables, `choose(p, 2)` pairwise tests are run simultaneously.
`prune_edges()` applies Benjamini-Hochberg false discovery rate control by
default. Holm and Bonferroni are also available.

**5. Unsigned weights.**
All four association metrics quantify strength, not direction. Two edges with
the same weight may reflect very different cell-level deviations from
independence. The signed standardised Pearson residual stored as `std_resid`
on modality-graph edges provides directional information.

### Additional caveats for the modality layer

Four caveats apply specifically to modality-level graphs.

**M1. Cross-variable edges only.**
Same-variable modalities are excluded by construction, because they are
mutually exclusive and a 2×2 indicator table between them has structural
zeros that make phi and residuals meaningless.

**M2. Unsigned weights; signed residuals stored separately.**
Edge weights are absolute association values, so attraction (modalities
co-occurring more than expected) and repulsion (less than expected) produce
edges of the same magnitude. The signed standardised Pearson residual is
stored as an edge attribute (`std_resid`) and can be used for interpretation.

**M3. Community detection groups modalities, not people.**
Communities are sets of co-associated category levels. They do not label
respondents. Assigning respondents to communities post-hoc is not supported
in this package because it conflates category co-association with latent-class
structure; for that, use `poLCA` or `FactoMineR::HCPC()`.

**M4. Rare modalities may fail to cluster meaningfully.**
Factor levels endorsed by few respondents have weak association with
everything else in the network and will typically end up as isolated
vertices (singleton communities) after pruning. This is not a bug but a
reflection of sparse data; consider collapsing rare categories before
building the graph if singleton isolates dominate the output.

---

## Example data

The package includes `survey_health`, a synthetic categorical dataset used
to illustrate the variable-level workflow.

```{r data}
str(survey_health)
```

For modality-level examples, the vignette uses the expanded `Titanic`
dataset because it is compact and easy to interpret.

---

## Building a variable-level graph

```{r build}
cg_full
summary(cg_full, top = 10)
```

---

## Similarity matrix vs graph

From v0.4.0 onward, two objects are separated clearly:

* **`assoc_similarity()`** returns a dense `p × p` matrix including pairs
  with zero association. Use this for matrix views and heatmaps.
* **A `catgraph` object** returns a sparse graph where zero-association
  pairs are absent edges. Use this for topology, centrality, and
  community detection.

```{r matrices}
S <- assoc_similarity(survey_health)
round(S, 3)

head(assoc_matrix(cg_full, format = "tidy"))
```

---

## Heatmap

```{r heatmap}
plot_heatmap(cg_full, title = "survey_health associations")
plot_heatmap(cg_full, show_sig = TRUE, title = "With significance stars")
```

---

## Graph visualisation

The `layout` argument controls graph arrangement.

```{r graph}
cg_pruned <- prune_edges(
  cg_full,
  min_weight = 0.05,
  max_p = 0.05,
  p_adjust = "BH"
)

plot(cg_pruned, layout = "fr",     title = "Fruchterman-Reingold")
plot(cg_pruned, layout = "kk",     title = "Kamada-Kawai")
plot(cg_pruned, layout = "circle", title = "Circular layout")
```

---

## Pruning with multiplicity control

```{r pruning}
# Raw p-values (legacy behavior; generally not recommended)
prune_edges(cg_full, max_p = 0.05, p_adjust = "none")

# BH FDR correction (default)
prune_edges(cg_full, max_p = 0.05, p_adjust = "BH")

# Holm-Bonferroni (more conservative)
prune_edges(cg_full, max_p = 0.05, p_adjust = "holm")
```

**Do not chain adjusted prunes.**
Calling `prune_edges()` with a p-value adjustment on a graph that has
already been adjusted re-scopes the multiplicity denominator to the
surviving edges, which is anti-conservative. If you need a different
adjustment method, rebuild the graph and prune once.

```{r chained-caveat, eval = FALSE}
cg_fresh <- catgraph(survey_health)
cg_final <- prune_edges(
  cg_fresh,
  min_weight = 0.1,
  max_p = 0.05,
  p_adjust = "holm"
)

cg_bh   <- prune_edges(cg_full, max_p = 0.5,  p_adjust = "BH")
cg_both <- prune_edges(cg_bh,   max_p = 0.05, p_adjust = "BH")
```

---

## Association paradigms

From v0.11.0, four association paradigms are available via the `method`
argument. All four produce edge weights in [0, 1] and are symmetric, making
them suitable for undirected graphs.

| `method` | Paradigm | Best for |
|---|---|---|
| `"cramers_v"` | Frequentist | Standard use, large samples |
| `"cramers_v_corrected"` | Frequentist | Sparse or small tables |
| `"nmi"` | Information-theoretic | Variables with unequal category counts |
| `"ami"` | Information-theoretic | Sparse tables, chance correction needed |
| `"bayesian_cramers_v"` | Bayesian | Small samples, rare categories |

```{r paradigms}
# All four on the same data — weights differ in scale but rankings agree
cg_cv  <- catgraph(survey_health, method = "cramers_v_corrected")
cg_nmi <- catgraph(survey_health, method = "nmi")
cg_ami <- catgraph(survey_health, method = "ami")
cg_bay <- catgraph(survey_health, method = "bayesian_cramers_v", alpha = 0.5)

# Print summaries
cg_cv
cg_nmi
```

```{r paradigms-compare}
# Compare edge weights across paradigms using assoc_similarity
S_cv  <- assoc_similarity(survey_health, method = "cramers_v_corrected")
S_nmi <- assoc_similarity(survey_health, method = "nmi")

ij <- which(upper.tri(S_cv), arr.ind = TRUE)
data.frame(
  var1 = rownames(S_cv)[ij[, 1]],
  var2 = colnames(S_cv)[ij[, 2]],
  CV   = round(S_cv[ij],  4),
  NMI  = round(S_nmi[ij], 4)
)
```

The key property to check is rank agreement across paradigms. High Spearman
correlation (ρ > 0.90) between paradigms indicates that structural conclusions
are robust to the choice of metric. Divergence reveals variable pairs where
the paradigm choice matters substantively — typically pairs involving variables
with skewed marginal distributions or sparse contingency tables.

```{r paradigms-rank}
# Rank correlation between CV and NMI — use similarity matrices
# so dimensions are always identical regardless of edge pruning
S_cv_full  <- assoc_similarity(survey_health, method = "cramers_v_corrected")
S_nmi_full <- assoc_similarity(survey_health, method = "nmi")

ij <- which(upper.tri(S_cv_full), arr.ind = TRUE)
w_cv_vec  <- S_cv_full[ij]
w_nmi_vec <- S_nmi_full[ij]

cat("Spearman rho (CV vs NMI):",
    round(cor(w_cv_vec, w_nmi_vec, method = "spearman"), 3), "\n")
```

For dense, large-n datasets the four paradigms will agree closely. For
datasets with rare categories or small samples, AMI and Bayesian Cramér's V
provide more stable estimates than their classical counterparts.

---

## Bootstrap confidence intervals

```{r ci}
ci <- bootstrap_ci(
  survey_health$smoking_status,
  survey_health$lung_disease,
  R = 1000,
  type = "percentile",
  seed = 1
)

cat(sprintf(
  "phi = %.3f  95%% CI [%.3f, %.3f]\n",
  ci$estimate, ci$ci_lower, ci$ci_upper
))

hist(
  ci$boot_dist,
  breaks = 30,
  main = "Bootstrap distribution",
  xlab = "phi-hat",
  col = "#EEEDFE",
  border = "#534AB7"
)

abline(v = ci$estimate, col = "#534AB7", lwd = 2)
abline(v = c(ci$ci_lower, ci$ci_upper), col = "#D85A30", lwd = 1.5, lty = 2)
```

---

## Community detection on variables

Remember that variable communities describe blocks of variables that
co-vary pairwise, not groups that are conditionally independent of the rest.

```{r clusters}
set.seed(42)

cg_clust <- detect_clusters(cg_full, method = "louvain")

cat("Clusters:", cg_clust$clustering$n_clusters, "\n")
cat("Modularity:", round(cg_clust$clustering$modularity, 3), "\n")

data.frame(
  variable = igraph::V(cg_clust$graph)$name,
  cluster  = igraph::V(cg_clust$graph)$cluster
)
```

---

## Modality-level network analysis

In addition to variable-level graphs, `catgraph` can construct a
**modality-level graph** in which each node is a modality (factor level)
from the original dataset. Edges are computed only across **different**
variables using the chosen association measure on binary indicator pairs.

This makes it possible to:

* detect communities among co-associated modalities,
* summarise the variable composition of each community,
* visualise modality graphs directly.

Communities are **groups of modalities (factor levels) that co-associate
across different variables** — not respondent segments. For
respondent-level segmentation, use the `poLCA` or `FactoMineR` packages.

### Build, prune, and cluster a modality graph

```{r modality-network-example}
df <- expand_table(Titanic)

mg <- build_modality_graph(df)
mg <- prune_modality_edges(mg, min_weight = 0.10)
mg <- cluster_modalities(mg)

mg
summary(mg)
```

### Visualise the modality graph

The graph can be coloured either by original variable or by detected
modality community.

```{r modality-network-plot, fig.width=8, fig.height=6}
plot(mg)
```

```{r modality-network-plot-cluster, fig.width=8, fig.height=6}
plot(mg, color_by = "cluster")
```

### Summarise modality communities

```{r modality-network-summary}
comm <- summarise_modality_communities(mg)

comm
summary(comm)
comm$variable_composition
```

### Interpretation note

The modality-network layer is a **category co-association map** in the
tradition of Multiple Correspondence Analysis (MCA) and two-mode
affiliation networks. Interpretation should focus on:

* which modalities co-associate across variables,
* which variables span multiple communities (potential bridges),
* community composition in terms of variable membership.

It is **not** a respondent-segmentation method. For latent-class
respondent models, see the `poLCA` package; for MCA plus hierarchical
clustering of respondents, see `FactoMineR::MCA()` and
`FactoMineR::HCPC()`.

---

## Complete workflows

### Variable-level workflow

```{r workflow-variable, eval = FALSE}
cg   <- catgraph(survey_health, method = "cramers_v_corrected")
plot_heatmap(cg, show_sig = TRUE)

cg   <- catgraph_ci(cg, R = 1000, seed = 42)
plot_heatmap(cg, show_values = TRUE, show_ci = TRUE)

cg_p <- prune_edges(cg, min_weight = 0.1, max_p = 0.05, p_adjust = "BH")
plot(cg_p, layout = "kk")

cg_p <- detect_clusters(cg_p, method = "louvain")
```

### Modality-level workflow

```{r workflow-modality, eval = FALSE}
df   <- expand_table(Titanic)

mg   <- build_modality_graph(df, method = "cramers_v_corrected")
mg   <- prune_modality_edges(mg, min_weight = 0.10)
mg   <- cluster_modalities(mg, method = "louvain")

plot(mg, color_by = "cluster")

comm <- summarise_modality_communities(mg)
summary(comm)
```

---

## References

* Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery rate.
  *Journal of the Royal Statistical Society: Series B*, 57(1), 289–300.
* Bergsma, W. (2013). A bias-correction for Cramér's V and Tschuprow's T.
  *Journal of the Korean Statistical Society*, 42(3), 323–328.
* Cover, T. M., & Thomas, J. A. (2006). *Elements of Information Theory*
  (2nd ed.). Wiley.
* Cramer, H. (1946). *Mathematical Methods of Statistics*. Princeton University Press.
* Efron, B., & Tibshirani, R. J. (1993). *An Introduction to the Bootstrap*. Chapman & Hall.
* Good, I. J. (1965). *The Estimation of Probabilities*. MIT Press.
* Holm, S. (1979). A simple sequentially rejective multiple test procedure.
  *Scandinavian Journal of Statistics*, 6(2), 65–70.
* Vinh, N. X., Epps, J., & Bailey, J. (2010). Information theoretic measures
  for clusterings comparison. *Journal of Machine Learning Research*, 11, 2837–2854.
